Topological Equivariant Artist Model

### 2021-03-23 paper/paper.tex
paper draft

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.
\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{mathrsfs} % math script fonts
\usepackage{tabulary}
\usepackage{subfig} %ieee does not like subfigure
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{cd} % commutative diagrams
\newtheorem{prop}{Proposition} %math?
\usepackage[switch]{lineno}
\usepackage{minted}
\setminted[python]{fontsize=\scriptsize, 
                   linenos,
                   numbersep=8pt,
                   autogobble, 
                   frame=lines,
                   framesep=3mm} 
\usepackage{notation} % move this later

@@

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization to improve consistency, reusability, and discoverability, so domain specific tool developers can build structure preserving visualization tools. The contribution of this work is 
\begin{enumerate}
  \item topology preserving relationship between data and graphic via continuous maps
  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
\end{enumerate}

\section{Related Work}
The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}. The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization. Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode. A fully expressive component is one that is equivariant since it has preserved all the structure in the data. 
Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations. Mackinlay's \textit{A Presentation Tool} (APT) introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}. This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. This same limitation is explicitly stated in the functional dependency model of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative. The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.

On the other hand, most information visualization tools are explicitly tuned to specific continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to tools influenced by APT, such as
Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.

@@

 \subsubsection{Sheafs}
 \label{sec:math:data:sheaf}
 The restriction maps of a sheaf describe how local sections $\iota^{*}\dsection$ over a neighborhood $U$ around \dbasepoint\ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}. This puts a continuous structure on local sections, which allows for defining a section over a subset in \dbase. The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$ 
 \begin{equation}
     \label{eq:sheaf}
     \begin{tikzcd}
         \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
         U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
 such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$. Gluing together $\iota^*\dsection$ sections is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.

@@

\subsection{Artist}
The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
\begin{equation}
    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
\end{equation}
that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}. Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps 
\begin{equation}
\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
\end{equation}
such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.

@@

\subsubsection{Visual Component Maps}
\label{sec:math:artist:nu}
 We define the visual transformers \vchannel\ 
\begin{equation}
  \label{eq:math:artist:nu}
  \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
\end{equation}
as the set of equivariant maps $\vchannel_i: \dsection_i \mapsto \vsection_i$. Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve. As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance. A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes
\begin{equation}
  \label{eq:math:artist:nu_commute}
\begin{tikzcd}
  \dtotal_i \arrow[r] \arrow[r, "\vchannel_i"] \arrow[d, "m_{\delement}"'] & \vtotal_i \arrow[d, "m_{\velement}"] \\
  \dtotal_i \arrow[r, "\vchannel_i"]                           & \vtotal_i               
\end{tikzcd}
\end{equation}
such that applying equivariant monoid actions to $\dtotal_{i}$ and $\vtotal_{i}$ preserves the map $\vchannel_{i}: \dtotal_{i}\rightarrow \vtotal_{i}$. In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$ and the actions of \monoid\ on $\dfiber_{i}$ cannot be assumed to be the same as the actions of $\monoid^{\prime}$ on \vfiber; therefore an equivariant $\vchannel_i$ must satisfy the constraint  
\begin{equation}
\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
\end{equation} 
such that $\varphi$ maps a monoid action on data to a monoid action on visual elements. However, we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$. We can compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi$ that takes \monoid\ to $\monoid^{\prime}$. This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$. Therefore, without a loss of generality, we can assume that an action of \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.

\begin{table}[H]
  %%\renewcommand{\arraystretch}{2}
  \begin{tabulary}{\columnwidth}{|llL|}\hline
      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
  \end{tabulary}
  \caption{}
  \label{tab:math:artist:nu}
\end{table}
The equivariance constraints on the Stevens measurement scales \cite{stevensTheoryScalesMeasurement1946} are listed in \autoref{tab:math:artist:nu}. These are derived from the group actions on the scales, as defined by Stevens.

\begin{figure}[htb]
  \centering
  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
  \caption{In the top two panels \vchannel\ maps to visual elements such that the relative order of the visual elements is the same as the data. In the bottom right panel, \textit{rain} is mapped to a visual element that is less than the visual element \textit{storm} is mapped to, which is not an allowed \vchannel. Changing the map so that \textit{storm} goes to the same visual element as \textit{rain} yields an order preserving map and is therefore a valid \vchannel.}
  \label{fig:math:artist:nu}
\end{figure}

The equivariance constraint for monoid actions is that relative order is preserved, as shown in the top two panels of \autoref{fig:math:artist:nu}. In the bottom right panel the ordering of \textit{rain} and \textit{storm} is swapped in the visual elements; in the bottom left panel, \textit{storm} is instead mapped to the same visual element as \textit{rain} so that the ordering is preserved.

\subsubsection{Visualization Assembly}
\label{sec:math:artist:q}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\columnwidth]{path_of_q}
  \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.} 
  \label{fig:math:artist:q}
\end{figure}
As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\ reduce the encodings into a graphic \gsection. The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$. We formalize the expectation that visualization generation functions parameterized in the same way should generate the same functions as the equivariant map $\vmark: \vsection \mapsto \gsection$. We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.  We cannot define monoid actions on all of $\Gamma(\gtotal)$ because there are graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
\begin{figure}[htb]
  \centering
  \includegraphics[width=\columnwidth]{diff_type_q.png}
  \caption{These two glyphs are generated by the same annulus \vmark\ function. The monoid action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
  \label{fig:math:artist:graphic}
\end{figure}
Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$
\begin{prop}
If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
\begin{equation}
\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
\label{eq:math:artist:q}
\end{equation}
then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$. If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection. 
\end{prop}

@@

Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out. When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
\begin{equation}
    \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
    \label{eq:math:artist:qhat}
\end{equation}
 is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint. Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function 
\begin{equation}
\label{eq:math:artist:qhat_q_s}
\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
\end{equation} 
where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.

Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase. In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$. This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$. The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.

>>>>

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.
\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{mathrsfs} % math script fonts
\usepackage{tabulary}
\usepackage[utf8]{inputenc}
\usepackage{subfig} %ieee does not like subfigure
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{cd} % commutative diagrams
\newtheorem{prop}{Proposition} %math?
\usepackage[switch]{lineno}
\usepackage{minted}
\setminted[python]{fontsize=\scriptsize, 
                   linenos,
                   numbersep=8pt,
                   autogobble, 
                   frame=lines,
                   framesep=3mm} 
\usepackage{notation} % move this later

@@

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}. Visualization tools allow practioners to automate these transformations, but most visualization tools are either explicitly tuned to specific continuity \cite{HeerSoftware2006} or define every interface relative to its expected data structure. The former approach is often too restrictive for a general purpose tool, while the latter often results in a library that can feel inconsistent or disjointed. Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation. The contribution of TEAM is 
\begin{enumerate}
  \item topology preserving relationship between data and graphic via continuous maps
  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
\end{enumerate}

\section{Related Work}
The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization. Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode. A fully expressive component is one that is equivariant since it has preserved all the structure in the data. Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations. Mackinlay's \textit{A Presentation Tool} (APT) introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}. This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. This same limitation is explicitly stated in the functional dependency model of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative. The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.

Much of visualization algorithm design is oriented around continuity, as described in Tory and Möller's taxonomy\cite{toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.

@@

 \subsubsection{Sheafs}
 \label{sec:math:data:sheaf}
A sheaf is a mathematical structure for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces. On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections $\iota^{*}\dsection$
 \begin{equation}
     \label{eq:sheaf}
     \begin{tikzcd}
         \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
         U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
which are sections of \dtotal\ pulled back over local neighborhood $U \subset \dtotal$ via the inclusion map \(\iota:\dtotal\rightarrow U\). The bookkeeping enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}

@@

\subsection{Artist}
The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
\begin{equation}
    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
    \label{eq:math:artist:artist}
\end{equation}
that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}. Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps 
\begin{equation}
\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
\end{equation}
such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.

@@

\subsubsection{Visual Component Maps}
\label{sec:math:artist:nu}
 We define the visual transformers \vchannel\ 
\begin{equation}
  \label{eq:math:artist:nu}
  \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
\end{equation}
as the set of equivariant maps $\vchannel_i: \dsection_i \mapsto \vsection_i$. Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve. As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance. A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes
\begin{equation}
  \label{eq:math:artist:nu_commute}
\begin{tikzcd}
  \dtotal_i \arrow[r] \arrow[r, "\vchannel_i"] \arrow[d, "m_{\delement}"'] & \vtotal_i \arrow[d, "m_{\velement}"] \\
  \dtotal_i \arrow[r, "\vchannel_i"]                           & \vtotal_i               
\end{tikzcd}
\end{equation}
such that applying equivariant monoid actions to $\dtotal_{i}$ and $\vtotal_{i}$ preserves the map $\vchannel_{i}: \dtotal_{i}\rightarrow \vtotal_{i}$. In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$ and the actions of \monoid\ on $\dfiber_{i}$ cannot be assumed to be the same as the actions of $\monoid^{\prime}$ on \vfiber; therefore an equivariant $\vchannel_i$ must satisfy the constraint  
\begin{equation}
\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
\label{eq:math:artist:nu:equivariance}
\end{equation} 
such that $\varphi$ maps a monoid action on data to a monoid action on visual elements. However, we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$. We can compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi$ that takes \monoid\ to $\monoid^{\prime}$. This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$. Therefore, without a loss of generality, we can assume that an action of \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.

\begin{figure}[htb]
  \centering
  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
  \caption{Equivariant \vchannel\ maps the weather data to corresponding symbols such that the relative order of the symbols is the same as the data as shown in the top two panels. In the bottom right panel the ordering of \textit{rain} and \textit{storm} is swapped in the symbols, meaning that the \vchannel\ is invalid; in the bottom left panel, \textit{storm} is instead mapped to the same symbol as \textit{rain} so that the ordering is preserved.}
  \label{fig:math:artist:nu}
\end{figure}

\begin{table}[H]
  \renewcommand{\arraystretch}{2}
  \begin{tabulary}{\columnwidth}{|llL|}\hline
      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
  \end{tabulary}
  \label{tab:math:artist:nu}
\end{table}
The equivariance constraints for the Steven measurement scales\cite{stevensTheoryScalesMeasurement1946} are derived from the mathematical group structure of the scales. For monoid actions, the equivariance constraint is that relative order is preserved, as shown in the top two panels of \autoref{fig:math:artist:nu}.

\subsubsection{Visualization Assembly}
\label{sec:math:artist:q}
\begin{figure}[htb]
  \centering
  \includegraphics[width=\columnwidth]{path_of_q}
  \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.} 
  \label{fig:math:artist:q}
\end{figure}
As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are analogous to a  map-reduce operation: data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal. The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$. We formalize the expectation that visualization generation functions parameterized in the same way should generate the same functions as the equivariant map $\vmark: \vsection \mapsto \gsection$. We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.  We cannot define monoid actions on all of $\Gamma(\gtotal)$ because there are graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
\begin{figure}[htb]
  \centering
  \includegraphics[width=\columnwidth]{diff_type_q.png}
  \caption{These two glyphs are generated by the same annulus \vmark\ function. The monoid action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
  \label{fig:math:artist:graphic}
\end{figure}
Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$
\begin{prop}
If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
\begin{equation}
\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
\label{eq:math:artist:q}
\end{equation}
then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$. If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection. 
\end{prop}

@@

Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out. When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
\begin{equation}
    \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
    \label{eq:math:artist:qhat}
\end{equation}
 is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint. Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function 
\begin{equation}
\label{eq:math:artist:qhat_q_s}
\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
\end{equation} 
where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase. Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields
\begin{equation}
\vmarkd(\vsection(k)) = \vmark(\vsectionpull)
\label{eq:math:artist:qhat_k}
\end{equation}
where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase. In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$. This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$. The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
